---
title: "US Crime Statistics"
author: "Andrew Brown"
date: "March 6, 2019"
output: html_document
---

#Table of Contents
1. [Data Acquisition](#DataAcq)
2. [Data Description](#DataDesc)
3. [Prominent Features and General Information](#PFGI)
4. [Interesting Trends](#IT)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
#This is all the code we need to set up all our data
#Read in csv file
crime <- read.csv("table-2.csv")
#Remove useless row on bottom
crime <- crime[-c(133),]
#Rename columns appropriately
names(crime) <- c("Region", "Year", "Population", "Violent Crime", "Violent Crime per 100k people", "Murder and nonnegligent manslaughter", "Murder and nonnegligent manslaughter per 100k people", "Rape revised definition", "Rape revised definition per 100k people", "Rape legacy definition", "Rape legacy definition per 100k people", "Robbery", "Robbery per 100k people", "Aggravated assault", "Aggravated assault per 100k people", "Property crime", "Property crime per 100k people", "Burglary", "Burglary per 100k people", "Larceny theft", "Larceny theft per 100k people", "Motor vehicle theft", "Motor vehicle theft per 100k people")
#Copy the row names so they each match for years 2015 and 2016
crime[seq(2,132,2),]$Region <- crime[seq(1,131,2),]$Region

#This next block of code cerates 10 sub-sets of the main data frame in order to make it easier to view certain parts at a time
#Creates a data frame with only the 2015 crime data
crime_2015 <- crime[seq(1,131,2),]
#Creates a data frame with only the 2016 crime data
crime_2016 <- crime[seq(2,132,2),]
#Creates a data fram with only the states crime data (Plus Puerto Rico and D.C.)
crime_states <- crime[-c(1,2,3,4,5,6,19,20,27,28,29,30,41,42,57,58,59,60,79,80,89,90,99,100,101,102,119,120),]
#Creates a data fram with only the larger regions crime data
crime_regions <- crime[c(1,2,3,4,5,6,19,20,27,28,29,30,41,42,57,58,59,60,79,80,89,90,99,100,101,102,119,120),]
#Creates separate data frames for the two years for both the states data frame and the regions one
crime_states_2015 <- crime_states[seq(1,103,2),]
crime_states_2016 <- crime_states[seq(2,104,2),]
crime_regions_2015 <- crime_regions[seq(1,27,2),]
crime_regions_2016 <- crime_regions[seq(2,28,2),]
#Creates a data frame called combined_crime_2015 that lists the 10 different crime rates for each state in 2015
combined_crime_2015 <- crime_states_2015[,-c(2,3,4,6,8,10,12,14,16,18,20,22)]
rownames(combined_crime_2015) <- combined_crime_2015$Region
combined_crime_2015 <- combined_crime_2015[,2:11]
#Creates a data frame called combined_crime_2016 that lists the 10 different crime rates for each state in 2016
combined_crime_2016 <- crime_states_2016[,-c(2,3,4,6,8,10,12,14,16,18,20,22)]
rownames(combined_crime_2016) <- combined_crime_2016$Region
combined_crime_2016 <- combined_crime_2016[,2:11]
```

## <a name="DataAcq"></a>Data Acquisition
This data set was found on the Federal Bureau of Investigation's website <https://ucr.fbi.gov/crime-in-the-u.s/2016/crime-in-the-u.s.-2016/topic-pages/tables/table-2>. The FBI used their records of crime in the United States to create the data set and uploaded it online. On the site I clicked the "download excel" button near the top left of the page. I used Microsoft Excel to edit the file and remove some unnecessary information and to convert the file to a csv I could use within R. I have included the csv file along with this report. Finally, I imported the csv file into R and used tools within R to make a few more changes to the data set and make it easier for me to read.

## <a name="DataDesc"></a>Data Description
This data set is a recording of the number of crimes committed in the United states by crime type and by region. It keeps track of 10 different types of crime: Violent Crime, Murder and Nonnegligent Manslaughter, Rape (revised definition), Rape (legacy definition), Robbery, Aggravated Assault, Property Crime, Burglary, Larceny Theft, and Motor Vehicle Theft. It records the total number of these crimes committed in one year in each of the 50 states (plus Puerto Rico and District of Columbia) as well as the total for the US and the total for 13 bigger regions of the US. The data was recorded for two consecutive years, 2015 and 2016.

## <a name="PFGI"></a>Prominent Features and General Information
There aren't any noticeable instances where a certain region has a lot of one type of crime but not a lot of another. Crime itself basically follows the same pattern regardless of location, with the same types of crimes being more prominent than others. Based on the total numbers of crimes committed for each individual crime type from 2015 and 2016 combined, the most committed crime was property crime and the least committed crime was murder and nonnegligent manslaughter. Each type of crime is shown on this bar graph.

```{r GRAPH1, echo=FALSE}
#Bar graph comparing the different types of crime by how many times each has been committed
#GRAPH 1
barplot(t(c(crime[1,seq(4,23,2)]+crime[2,seq(4,23,2)])))
```

The bars along the X-axis (from left to right respectively) are Violent Crime, Murder and Nonnegligent Manslaughter, Rape (revised definition), Rape (legacy definition), Robbery, Aggravated Assault, Property Crime, Burglary, Larceny Theft, and Motor Vehicle Theft. The Y-axis represents number of crimes committed.

## <a name="IT"></a>Interesting Trends
### <a name="I"></a>I.
This graph shows the rape (revised definition) rate from 2016 for each state and Puerto Rico and District of Columbia.
```{r GRAPH2, echo=FALSE}
#Plot for Alaska having the highest rape rate by a lot compared to other states
#GRAPH 2
barplot(crime_states_2016$`Rape revised definition per 100k people`, col="orange")
```

Alaska has the highest rape rate at 141.9 per 100,000 people. DC has the 2nd highest at 78.1, and New Mexico is 3rd with 73.3. I found it fascinating that Alaska's was so much higher than any other state, but aside from Alaska, most of all the other states form a fairly tight distribution without much variance.

### <a name="II"></a>II.
These graphs show a comparison of the different types of crimes from 2015 and 2016.
```{r GRAPH 3 & 4, echo=FALSE}
#Process of creating the data frame to be used to compare the two years
#First part for 2015
crime_total_2015 <- rbind(combined_crime_2015, c(sum(combined_crime_2015$`Violent Crime per 100k people`), sum(combined_crime_2015$`Murder and nonnegligent manslaughter per 100k people`), sum(combined_crime_2015$`Rape revised definition per 100k people`), sum(combined_crime_2015$`Rape legacy definition per 100k people`), sum(combined_crime_2015$`Robbery per 100k people`), sum(combined_crime_2015$`Aggravated assault per 100k people`), sum(combined_crime_2015$`Property crime per 100k people`), sum(combined_crime_2015$`Burglary per 100k people`), sum(combined_crime_2015$`Larceny theft per 100k people`), sum(combined_crime_2015$`Motor vehicle theft per 100k people`)))
rownames(crime_total_2015)[53] <- "2015 totals"
crime_total_2015<- crime_total_2015[53,1:10]
#Second part for 2016
crime_total_2016 <- rbind(combined_crime_2016, c(sum(combined_crime_2016$`Violent Crime per 100k people`), sum(combined_crime_2016$`Murder and nonnegligent manslaughter per 100k people`), sum(combined_crime_2016$`Rape revised definition per 100k people`), sum(combined_crime_2016$`Rape legacy definition per 100k people`), sum(combined_crime_2016$`Robbery per 100k people`), sum(combined_crime_2016$`Aggravated assault per 100k people`), sum(combined_crime_2016$`Property crime per 100k people`), sum(combined_crime_2016$`Burglary per 100k people`), sum(combined_crime_2016$`Larceny theft per 100k people`), sum(combined_crime_2016$`Motor vehicle theft per 100k people`)))
rownames(crime_total_2016)[53] <- "2016 totals"
crime_total_2016 <- crime_total_2016[53,1:10]
#Putting the parts together
crime_total <- rbind(crime_total_2015, crime_total_2016)
#GRAPH 3 & 4 now the same because of par
par(mfrow=c(1,2))
barplot(t(crime_total[1,]), col="red", beside=TRUE)
barplot(t(crime_total[2,]), col="blue", beside=TRUE)
```

The Y-axis shows the totals of the crime rates for each type of crime. This means that the height of the bar represents the sum of every state's crime rate for that specific crime. These two graphs are almost identical, which shows that from 2015 to 2016 crime really didn't go up or down, it was just a normal small fluctuation.

### <a name="III"></a>III.
This graph represents each of the 50 state's (plus DC and Puerto Rico) total crime rate for 2016.
```{r GRAPH 5, echo=FALSE}
#Bar graph of the combined crime rates for 2016 by state
#GRAPH HERE 5
par(mfrow=c(1,1))
barplot(t(combined_crime_2016))
```

The entire bar is the total of all the crime rates, but the different colors within each bar represent the different types of crime. DC is the tallest bar, which means it has the highest overall crime rate. DC also happens to have the highest crime rate in violent crime, murder and nonnegligent manslaughter, robbery, aggravated assault, property crime, and larceny theft. The second highest overall crime rate is New Mexico, and Alaska is 3rd. Puerto Rico has the lowest crime rate, New Hampshire has the 2nd lowest, and Maine the third lowest. It's difficult to make a conjecture about crime rates based on this information, DC has a very compact population, but Alaska does not for example. But they are both very high in crime rates.

### <a name="IV"></a>IV.
These graphs show the correlation between burglary and all other types of crime.
```{r GRAPH 6, echo=FALSE}
#This block creates the 'pairs plot' for burglary to show that it doesn't really corrolate to other types of crime
#GRAPH 6
par(mfrow=c(3,3))
plot(combined_crime_2016[,8], combined_crime_2016[,1], main="Burglary by Violent Crime")
plot(combined_crime_2016[,8], combined_crime_2016[,2], main="Burglary by Murder")
plot(combined_crime_2016[,8], combined_crime_2016[,3], main="Burglary by Rape (RD)")
plot(combined_crime_2016[,8], combined_crime_2016[,4], main="Burglary by Rape (LD)")
plot(combined_crime_2016[,8], combined_crime_2016[,5], main="Burglary by Robbery")
plot(combined_crime_2016[,8], combined_crime_2016[,6], main="Burglary by Aggravated Assault")
plot(combined_crime_2016[,8], combined_crime_2016[,7], main="Burglary by Property Crime")
plot(combined_crime_2016[,8], combined_crime_2016[,9], main="Burglary by Larceny Theft")
plot(combined_crime_2016[,8], combined_crime_2016[,10], main="Burglary by Motor Vehicle Theft")
```

Each point on a scatter plot is one of the 50 states (plus DC and Puerto Rico) and its position on the X-axis represents its burglary rate, while its position on the Y-axis represents the crime rate for the other category, both for 2016. There are no strong or noticeable clusters of the points, so burglary does not appear to be correlated with other types of crimes, meaning if a region has a lot of burglary or not very much that won't really allow you to predict if their other types of crime are high or low or not. Creating similar graphs for all the other crime types showed that other crimes had more correlation with each other, and burglary had the least correlation with the others.

### <a name="V"></a>V.
This graph shows the correlation between aggravated assault and violent crime.
```{r GRAPH 7, echo=FALSE}
#Graph for aggravated assault vs violent crime becuase they have the strongest correlation among the crime types as found using a pairs plot
#GRAPH 7
par(mfrow=c(1,1))
plot(combined_crime_2016$`Aggravated assault per 100k people`, combined_crime_2016$`Violent Crime per 100k people`)
```

These two types of crimes seem to be very closely related to one another. Each point on the scatter plot represents one of the 50 states (plus DC and Puerto Rico), its position on the X-axis represents the aggravated assault crime rate, and its position on the Y-axis represents the violent crime rate, both for 2016. Notice that the points closely follow a linear pattern, this means that they do correlate with each other and each one could be a good predicter of the other. For example if a state has a low violent crime rate, it's likely they also have a low aggravated assault rate. To see just how strong this correlation was I performed a t test (as shown below) to see that the p value was 0.00001371! This is an extremely small value which shows that the correlation does exist.
```{r t test}
#This gives the p-value for the correlation of aggravated assault and violent crime
t.test(combined_crime_2016$`Aggravated assault per 100k people`, combined_crime_2016$`Violent Crime per 100k people`)
```